Measuring the double-drift illusion with hand tracking

### 2021-02-26 Manuscript.Rmd
minor updates to the manuscript

*: Corresponding author: \
Marius ’t Hart\
Centre for Vision Research\
York University\
Toronto, ON M3J1P3\
Canada\
thartbm@gmail.com\

@@

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background [@Gurnsey2012; @Kwon2015; @Cavanagh2019]. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position. If the gabor itself is not moving [@DeValois1991], the shift saturates quickly [40 _ms_: @Chung2007; 100 _ms_: @Arnold2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues well beyond 100 ms. The perceived position conintues to move away from its physical position up to several times the size of the gabor. The strength of the illusion, or the deviation of perceived path of the Gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015]; @Cavanagh2019. In our first experiment, participants track their percept with a stylus as the gabor moves back and forth on its trajectory (Fig. 1) with 5 speeds of the internal motion. We analyze the angle of the tracking to determine if manual response is susceptible to the illusion.

@@

The durations of each pass in the online tracking experiment was 2 seconds but the manual traces mostly showed a linear trajectory without saturation or reset of the types shown in Figure 4 on the left below. The continuation of the accumulation may be a result of the active tracking – either the extra attention required to track increased the extent of the accumulation or possibly the temporal demands of the tracking caused the participants to average their traces over any breaks. To address this, we switched to an offline recording of the perceived path.

```{r fig4b, echo=F, fig.height=3.2, fig.width=5.62, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 4: Resets.** _Left_: At some point, the perceived position will stop moving further away from its actual path. For some people this takes the form of a ‘jump’ back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combnation of the two. _Top right_: if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate. _Bottom right_: if they occur instead after some period of time, they should have a constant Y coordinate."}
library(magick)
Fig4b <- image_read_svg('doc/methods_fig_2.svg')
par(mai=c(0,0,0,0))
plot(Fig4b)
```

@@

```{r fig4, echo=FALSE, message=FALSE, fig.width=10, fig.height=4, fig.cap="\\label{fig:fig4}**Figure 4. Illusion Strength and Example Trajectoriesin Experiment 2.** _Left_: For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show the distribution of angles for points halfway between the start point and detected resets; our measure of illusion strength. Dots below the distribution are individual participants' average illusion strengths. Gray lines indicate no illusion. _Right_: One exemplary participant's trajectories with very clear 'jump' resets at the purple dots. (Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories without resets.) Other participants' data was not this clear."}
plotExampleData()
```

@@

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 7 (left), we can see that the reset points vary more along the Y-axis than the X-axis. That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff). In fact, the standard deviation of reset point X-coordinates is about 21.9% of the standard deviation of reset point Y-coordinates. This already indicates that a spatial limit explains about 4 times more of the variance in reset points than a temporal limit.

@@

The limits that optimally fit the model to the data are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit. While the AIC of the model (-209.8) is marginally lower than that of the model with the single spatial limit, but it does not explain the data better ( _p_=.53). This means that a simple spatial limit accounts for most of the data. It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, as it may not follow a normal distribution, which we do not account for here. However, regardless of whether or not a temporal limit adds anything, a spatial limit appears to explain the majority of spontaneous double-drift resets.

@@

While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead. An as of yet unpublished study [@Liu2019] obtained data on the PSE of a double-drift stimulus at the end of trajectories of various lengths traveled at different speeds. These PSEs were predicted very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7% / _s_). Interestingly, this temporal constraint fully contradicts the current findings. The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred. If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then we should see a similar pattern in the averaged trajectories in our re-tracing experiment. However, these averaged trajectories reach a plateau within ~1 _s_ after which resets and illusory position shift seem to balance each other. This means that we can't replicate the pattern found with the PSE approach. There may be several reasons for this. First, our data may simply be too noisy, or the effects may be highly dependent on participants so that both studies may suffer from sampling bias. We can't reolve this contradiction here and leave that for future work.

>>>>

*: Corresponding author: \
Marius ’t Hart\
Centre for Vision Research\
York University\
Toronto, ON M3J1P3\
Canada\
[thartbm@gmail.com](thartbm@gmail.com)\

@@

The strength of the double-drift illusion probably arises because of the very poor location information for a gabor pattern in the periphery when it has the same mean luminance as the background [@Gurnsey2012; @Kwon2015; @Cavanagh2019]. With poor positional certainty, the motion information contributes to the location estimate, generating a shift in perceived position. If the gabor itself is not moving [@DeValois1991], the shift saturates quickly [40 _ms_: @Chung2007; 100 _ms_: @Arnold2007], perhaps because the accumulating evidence of actual location of the stationary gabor patch is sufficient to anchor its position. However, when the gabor itself also moves as in the double-drift stimulus, there is no stable location information to anchor the position estimate and the build-up continues well beyond 100 ms. The perceived location conintues to move away from its physical position by up to several times the size of the gabor. The strength of the illusion, the deviation of perceived path of the Gabor from its true path, depends on the speed of both the internal and external motion [@Tse2006; @Shapiro2010; @Kwon2015; @Cavanagh2019]. In our first experiment, participants track their percept with a stylus as the gabor moves back and forth on its trajectory (Fig. 1) with 5 speeds of the internal motion. We analyze the angle of the tracking to determine if the manual response is susceptible to the illusion.

@@

The durations of each pass in the online tracking experiment was 2 seconds but the manual traces mostly showed a linear trajectory without saturation or reset of the types shown in Figure 4 on the left below. The continuation of the accumulation may be a result of the active tracking – either the extra attention required to track increased the extent of the accumulation or possibly the temporal demands of the tracking caused the participants to average their traces over any breaks. To address this, we switched to a delayed, offline recording of the perceived path.

```{r fig4b, echo=F, fig.height=3.2, fig.width=5.62, message=FALSE, error=FALSE, warning=FALSE, fig.cap="**Figure 4: Resets.** **Left:** At some point, the perceived position will stop moving further away from its actual path. For some people this takes the form of a ‘jump’ back toward the true position, for others it means the illusory position remains at a fixed offset moving parallel to the real path, or some combination of the two. **Top right:** if these resets are due to a spatial limit on the size of the illusory position shift away from the true vertical path, the locations of the resets should have a constant X coordinate. **Bottom right:** if they occur instead after some period of time, they should have a constant Y coordinate."}
library(magick)
Fig4b <- image_read_svg('doc/methods_fig_2.svg')
par(mai=c(0,0,0,0))
plot(Fig4b)
```

@@

```{r fig5, echo=FALSE, message=FALSE, fig.width=10, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5. Illusion Strength and Example Trajectoriesin Experiment 2.** _Left_: For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show the distribution of angles for points halfway between the start point and detected resets; our measure of illusion strength. Dots below the distribution are individual participants' average illusion strengths. Gray lines indicate no illusion. _Right_: One exemplary participant's trajectories with very clear 'jump' resets at the purple dots. (Purple lines: trajectories up to resets, gray lines: trajectories after resets, blue lines: trajectories without resets.) Other participants' data was not this clear."}
plotExampleData()
```

@@

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 6 (left), we can see that the reset points vary more along the Y-axis than the X-axis. That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff). In fact, the standard deviation of reset point X-coordinates is about 21.9% of the standard deviation of reset point Y-coordinates. This already indicates that a spatial limit explains about 4 times more of the variance in reset points than a temporal limit.

@@

The limits that optimally fit the model to the data are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit. While the AIC of the model (-209.8) is marginally lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.53). This means that a simple spatial limit accounts for most of the data. It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, as it may not follow a normal distribution, which we do not account for here. However, regardless of whether or not a temporal limit adds anything, a spatial limit appears to explain the majority of spontaneous double-drift resets.

@@

While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead. An as of yet unpublished study [@Liu2019] obtained data on the PSE of a double-drift stimulus at the end of trajectories of various lengths traveled at different speeds. These PSEs were predicted very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7% / _s_). Interestingly, this temporal constraint fully contradicts the current findings. The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred. If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then we should see a similar pattern in the averaged trajectories in our re-tracing experiment. However, these averaged trajectories reach a plateau within ~1 _s_ after which resets and illusory position shift seem to balance each other. This means that we can't replicate the pattern found with the PSE approach. There may be several reasons for this. Our data may simply be too noisy, or the effects may be highly dependent on participants so that both studies may suffer from sampling bias. But primarily, the PSE-based approach relies on all participants to exhibit full jump-resets, whereas the approach presented here is also valid for weaker resets. Nevertheless, we can't reolve this contradiction here and leave that for future work.
